---
title: "Olympic Medal Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = TRUE,
  message     = FALSE,
  warning     = FALSE,
  fig.width   = 8,
  fig.height  = 5,
  out.width   = "100%",
  fig.align   = "center"
)
library(tidyverse)
library(countrycode)
library(WDI)
library(maps)
library(viridis)
library(corrplot)
````

# Introduction

This analysis uses the TidyTuesday “olympics” dataset and World Bank 2016 indicators to:

1. **Aggregate** all‐time medal counts by NOC
2. **Augment** with population and GDP per capita
3. **Visualize** global performance and relationships
4. **Model** through correlation, linear regression, and clustering

# 1. Data Loading & Preparation

```{r load-data}
# 1.1 Fetch TidyTuesday Olympics data
url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv"
olympics <- read_csv(url, show_col_types = FALSE)

# 1.2 Aggregate medals by NOC
medals <- olympics %>%
  filter(!is.na(noc)) %>%
  group_by(noc) %>%
  summarise(
    gold   = sum(medal == "Gold",   na.rm = TRUE),
    silver = sum(medal == "Silver", na.rm = TRUE),
    bronze = sum(medal == "Bronze", na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(total = gold + silver + bronze)

# 1.3 Fetch 2016 population & GDP per capita
wb_ind <- c(pop = "SP.POP.TOTL", gdp_pc = "NY.GDP.PCAP.KD")
wb_raw <- WDI(country = unique(medals$noc),
              indicator = wb_ind,
              start = 2016, end = 2016,
              extra = FALSE) %>%
  select(iso2c, pop, gdp_pc)

# 1.4 Harmonize codes & join
wb <- wb_raw %>%
  mutate(noc = countrycode(iso2c, "iso2c", "iso3c")) %>%
  select(noc, population = pop, gdp_pc)

medals <- medals %>%
  left_join(wb, by = "noc") %>%
  mutate(country = countrycode(noc, "iso3c", "country.name")) %>%
  drop_na(country)

# Show a preview
knitr::kable(head(medals), caption = "Medals & World Bank Indicators")
```

# 2. Top Performers

## 2.1 Top 10 by Total Medals

```{r top10-total, fig.cap="Top 10 NOCs by All-Time Total Medals"}
top10 <- medals %>%
  arrange(desc(total)) %>%
  slice(1:10)

ggplot(top10, aes(reorder(country, total), total, fill = country)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Top 10 NOCs by Total Medals",
    x     = NULL, y = "Total Medals"
  ) +
  theme_minimal()
```

## 2.2 Top 10 by Medals per Million People

```{r top10-percap, fig.cap="Top 10 NOCs by Medals per Million Inhabitants"}
medals <- medals %>%
  mutate(medals_per_million = total / (population / 1e6))

top10_pc <- medals %>%
  arrange(desc(medals_per_million)) %>%
  slice(1:10)

ggplot(top10_pc, aes(reorder(country, medals_per_million), medals_per_million, fill = country)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Top 10 NOCs by Medals per Million People (2016)",
    x     = NULL, y = "Medals per Million"
  ) +
  theme_minimal()
```

# 3. Global Map

```{r choropleth, fig.cap="World Map of Total Medals by Country"}
world <- map_data("world") %>%
  mutate(iso3 = countrycode(region, "country.name", "iso3c"))

map_df <- world %>%
  left_join(medals, by = c("iso3" = "noc"))

ggplot(map_df, aes(long, lat, group = group, fill = total)) +
  geom_polygon(color = "gray80", size = 0.1) +
  scale_fill_viridis(
    option    = "magma",
    name      = "Total Medals",
    na.value  = "gray90"
  ) +
  labs(title = "All-Time Olympic Medals by Country") +
  theme_void()
```

# 4. Medal‐Type Distribution

```{r distribution, fig.cap="Overall Medal-Type Distribution"}
overall <- medals %>%
  summarise(across(gold:bronze, sum, na.rm = TRUE)) %>%
  pivot_longer(everything(), names_to = "Medal", values_to = "Count")

ggplot(overall, aes(x = "", y = Count, fill = Medal)) +
  geom_col(width = 1) +
  coord_polar("y") +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "Overall Olympic Medal Distribution") +
  theme_void()
```

# 5. Distributions & Relationships

## 5.1 Histogram of Total Medals

```{r hist-total, fig.cap="Histogram of Total Medals Across NOCs"}
ggplot(medals, aes(total)) +
  geom_histogram(binwidth = 50, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Total Medals",
    x     = "Total Medals",
    y     = "Frequency"
  ) +
  theme_minimal()
```

## 5.2 Gold vs. Silver

```{r scatter-gs, fig.cap="Scatter Plot: Gold vs Silver Medals"}
ggplot(medals, aes(gold, silver)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(
    title = "Gold vs Silver Medals",
    x     = "Gold",
    y     = "Silver"
  ) +
  theme_minimal()
```

## 5.3 Total Medals vs. Population

```{r scatter-pop, fig.cap="Total Medals vs Population (log scale)"}
ggplot(medals, aes(population, total)) +
  geom_point(alpha = 0.6) +
  scale_x_log10(labels = scales::comma) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  labs(
    title = "Total Medals vs Population (2016)",
    x     = "Population (log scale)",
    y     = "Total Medals"
  ) +
  theme_minimal()
```

# 6. Correlation Analysis

```{r correlation, fig.cap="Correlation Matrix of Medal Counts & Efficiency"}
corr_vars <- medals %>%
  select(gold, silver, bronze, total, medals_per_million)

corr_mtx <- cor(corr_vars, use = "pairwise.complete.obs")
corrplot(
  corr_mtx,
  method = "color",
  type   = "upper",
  tl.col = "black",
  tl.cex = 0.8,
  title  = "Correlation Heatmap"
)
```

# 7. Regression Modeling

```{r regression}
mod <- lm(total ~ population + gdp_pc, data = medals)
summary(mod)
```

```{r diag-plot, fig.cap="Residuals vs Fitted for Regression Model"}
plot(mod, which = 1)
```

# 8. K‐Means Clustering

```{r clustering, fig.cap="PCA Scatter with k-means Clusters (k=4)"}
# prepare & scale
cl_data <- scale(corr_vars)
set.seed(42)
km <- kmeans(cl_data, centers = 4, nstart = 25)

# PCA for 2D projection
pca <- prcomp(cl_data)
scores <- as_tibble(pca$x[,1:2]) %>%
  bind_cols(country = medals$country, cluster = factor(km$cluster))

ggplot(scores, aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.8) +
  labs(
    title = "PCA of NOC Profiles with k-means Clusters",
    x     = "PC1", y = "PC2"
  ) +
  theme_minimal() +
  scale_color_viridis_d()
```

# 9. Conclusions

* **Raw powerhouses** (USA, URS, GER…) dominate total counts.
* **Efficiency leaders** (small nations with high medals per capita) emerge in per-million ranking.
* **Population & GDP** are statistically significant predictors of raw medal counts.
* **Clusters** reveal distinct country profiles:

  1. Large delegations
  2. Wealthy mid-sized teams
  3. Efficiency specialists
  4. Low-output countries

---

*Analysis performed using R 4.3, tidyverse, countrycode, WDI, maps, viridis & corrplot.*
